---
title: "Changing demographic rates"
author: "Eric J. Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Changing demographic rates}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE,
  comment = "#>"
)
library(kwdemog)
library(dplyr)
library(mgcv)
library(ggplot2)
data(orca)
year.end = as.numeric(substr(Sys.Date(),1,4))
refit_models = FALSE
report_start = year.end-5
# format whale names just for consistency
format_names = function(x) {
  for(i in 1:2) {
  indx = which(substr(x,2,2) == "0")
  x[indx] = paste0(substr(x[indx],1,1), substr(x[indx],3,nchar(x[indx])))
  }
  return(x)
}

expanded = expand(orca)

expanded$animal = format_names(expanded$animal)
expanded$pod = format_names(expanded$pod)
expanded$matriline = format_names(expanded$matriline)
expanded$mom = format_names(expanded$mom)
expanded$dad = format_names(expanded$dad)

#report_dir = "projections/"
whaleData = expanded

```

### Approach  

We can calculate trends in demographic rates (survival, fecundity) for this dataset, using a framework similar to that described in the PFMC Working Group reports.

### Survival

```{r}
whaleData[which(whaleData$animal %in% c("L095","L098","L112") & whaleData$alive==0),] = NA

# filter out animals born > 1960
whaleData = dplyr::filter(whaleData, birth > 1960)
# fill in missing sexes randomly
set.seed(123)
whaleData$sexF1M2[which(whaleData$sexF1M2==0)] = sample(1:2, size=length(which(whaleData$sexF1M2==0)), replace=T)

fit <- gam(alive ~ s(year) + age+I(age^2)+sexF1M2,data=dplyr::filter(whaleData,includeSurv==1),
           family="binomial")

newdat = expand.grid(year =1976:max(whaleData$year),
                     sexF1M2=c(1,2),
                     age=20)
newdat$Sex = c("Female","Male")[newdat$sexF1M2]
newdat$pred = predict(fit, newdat,type="link")
newdat$pred_se = predict(fit, newdat, se.fit=TRUE,type="link")$se.fit

newdat$survival = plogis(newdat$pred)
newdat$lo = plogis(newdat$pred-1.96*newdat$pred_se)
newdat$hi = plogis(newdat$pred+1.96*newdat$pred_se)

newdat = dplyr::select(newdat, -pred, -pred_se)
```

Figure 1. Estimated survival rates for 20-year old males and females in the SRKW population. Ribbons represent 95% CIs. 

```{r}
ggplot(newdat, aes(year,survival,col=age,fill=age)) + 
  geom_ribbon(aes(ymin=lo,ymax=hi),alpha=0.3,col=NA) + 
  geom_line() +
  scale_color_viridis_c(end=0.8) + 
  scale_fill_viridis_c(end=0.8) + 
  facet_wrap(~Sex,nrow=2,scale="free_y")+
  theme_bw() + 
  xlab("Year") + ylab("Fecundity") + 
  theme(legend.position = "none") + 
  theme(strip.background =element_rect(fill="white"))
```

### Fecundity

```{r}
whaleData = dplyr::filter(whaleData, sexF1M2==1, 
                          includeFec==1, 
                          age >= 10,
                          age<=43,
                          alive == 1,
                          !is.na(gave_birth))

fit <- gam(gave_birth ~ s(year) + age+I(age^2),data=whaleData,
           family="binomial")

newdat_f = expand.grid(year =1976:max(whaleData$year),
                     age=20)
newdat_f$pred = predict(fit, newdat_f,type="link")
newdat_f$pred_se = predict(fit, newdat_f, se.fit=TRUE,type="link")$se.fit

newdat_f $fecundity = plogis(newdat_f$pred)
newdat_f $lo = plogis(newdat_f$pred-1.96*newdat_f$pred_se)
newdat_f $hi = plogis(newdat_f$pred+1.96*newdat_f$pred_se)

newdat_f = dplyr::select(newdat_f, -pred, -pred_se)
```


Figure 2. Estimated fecundity rates for 20-year females in the SRKW population. Ribbons represent 95% CIs. 
```{r}
ggplot(newdat_f, aes(year,fecundity,col=age,fill=age)) + 
  geom_ribbon(aes(ymin=lo,ymax=hi),alpha=0.3,col=NA) + 
  geom_line() +
  scale_color_viridis_c(end=0.8) + 
  scale_fill_viridis_c(end=0.8) + 
  theme_bw() + 
  xlab("Year") + ylab("Fecundity") + 
  theme(legend.position = "none") + 
  theme(strip.background =element_rect(fill="white"))
```

Table 1. Estimated time-varying survival rates for 20-year old female and male killer whales from the SRKW population. The 'lo' and 'hi' values represent 95% confidence intervals.  

```{r}
newdat$survival = round(newdat$survival,3)
newdat$lo = round(newdat$lo,3)
newdat$hi = round(newdat$hi,3)
knitr::kable(newdat,digits = 3)
```

Table 2. Estimated time-varying fecundity rates for 20-year old female killer whales from the SRKW population. The 'lo' and 'hi' values represent 95% confidence intervals.  

```{r}
newdat_f$fecundity = round(newdat_f$fecundity,3)
newdat_f$lo = round(newdat_f$lo,3)
newdat_f$hi = round(newdat_f$hi,3)
knitr::kable(newdat_f,digits = 3)
```
